To be FAIR: Theory Specification Needs an Update

### 2025-06-19 action_letter.Rmd
update 2025-06-19 09:59:24.514362

Thank you for considering our manuscript for publication at _`r rmarkdown::metadata$journal`_.
We appreciate the feedback that you and the reviewers have provided.
In the following itemised list we respond to each comment point by point.

@@

Upload tables separately: tab_meehl.csv
Upload figures separately: ec_visio.pdf, git_branches.png, git_compare_short.png

@@

This does not directly ask for a change, but summarize relevant changes here.

<!-- AB: -->
1.	I agree that striking the balance between the conceptual and technical details is the very challenge of this paper. I think we should go a route in the revision in which we try to explain technical details in even simpler language. For example, the github reference to our code could read like this: "Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #XXXXX." And maybe there is a way to move the more technical details to a section that can be skipped without compromising the conceptual/opinion aspects. @wilkinsonFAIRGuidingPrinciples2016

@@

As methodologists, our focus in this paper is on "procedures", rather than policies or documentation. Our goal is to position FAIR theory as a communication protocol: a set of standards, grounded in philosophy of science and implemented in existing infrastructure, that facilitates theory specification, reuse, and collaborative refinement.

@@

Compare FAIR theory to preregistration: it, too, was introduced first as a good idea; now - a decade later - empirical studies are being conducted to investigate its effects.

@@

@@

@@

We respectfully disagree; open data - one example of modular publishing - has increased dramatically in recent decades, up from ~7% of articles in 2000–2009 to ~50% of psychological articles in 2010-2019, and more (~80%) in other disciplines.
Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in psychology:

@@

<font color = "red"> TO DO </font>

@@

on behalf of all co-authors, I would like to express our sincere appreciation for your thoughtful comments. This is my (Caspar's) favorite review to date, and I especially value your perspective as a seasoned scholar/theoretician. All authors unanimously agreed that it would be a pity to deprive our readers of an opportunity to view FAIR theory through the lens of your lived experience - but we were not sure how to incorporate it. We invite your, and the Editor's, advice on how to best go about it.

@@

We think that this comment signals a need to better explain the important distinctions between these two practices.
To address this comment, we have included the following section in the revision:

@@

There are now three worked examples for actual theories in the package vignettes:
[For Self-Determination Theory](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html), the [Dunning-Kruger Effect](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html), and [Morris' Tripartite Model](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html),
which we reference in the text.

@@

That said, we are not sure that we are best positioned to engage with these issues, nor that the present paper would be the best place for such an argument.
As methodologists, our primary aim in this paper is to articulate the affordances of FAIR theory:
a framework that supports better theory specification, sharing, and reuse.
We believe our contribution will be most effective if it remains sharply focused on the practical infrastructure and its scientific use cases.

>>>>

Thank you for considering our manuscript for publication at _`r rmarkdown::metadata$journal`_.
We appreciate the feedback that you and the reviewers have provided.
In this action letter, we respond to each comment point by point.

@@

We have 
Upload tables separately: tab_meehl.csv
Upload figures separately: ec_visio.pdf, git_branches.png, git_compare_short.png

@@

We understand that striking the balance between the conceptual and technical details is the very challenge of this paper. In the revision we have tried to explain technical details in even simpler language.
The most important change we have made, as requested by Reviewer #2, is to move the tutorial parts of the paper into package vignettes.
We still leave a conceptual overview of the tutorial in simpler language because we do want readers of the paper to come away with some idea of what making a theory FAIR entails, and because these general principles are not tied to the R-pacakge and our specific implementation (they could be implemented using other infrastructures).

With regard to your specific example of technical language that might elude the readership, the GitHub reference in the author note now reads: "Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #XXXXX."

@@

To clarify our perspective: As methodologists, our focus is on "procedures", rather than policies or documentation. Our goal in this paper is to position FAIR theory as a set of standards, grounded in philosophy of science and implemented in existing infrastructure, that facilitates theory specification, reuse, and collaborative refinement.

@@

Compare FAIR theory to preregistration: it, too, was introduced first as a good idea.
Only now - a decade later - are empirical studies are being conducted to investigate its effects.on scientific practices.

@@

**DO THIS**

@@

**DO THIS, ANDREAS?**

@@

We respectfully disagree; open data - one example of modular publishing - has increased dramatically in recent decades, in psychology the percentage of articles that share data has increased from ~7% in 2000–2009 to ~50% in 2010-2019. In other disciplines, the increase has been even larger (up to ~80%).
Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in psychology:

@@

We apologize for the oversight, and have fixed these hyperlinks.

@@

on behalf of all co-authors, let me express our sincere appreciation for your thoughtful comments. This is my (Caspar's) favorite review to date, and I especially value your perspective as a seasoned scholar/theoretician. We have attempted to address your comments in the Revision, as described below. In addition, all authors agreed that it would be a pity to deprive our readers of an opportunity to view FAIR theory through the lens of your lived experience - but we were not entirely sure how to incorporate it. We invite your, and the Editor's, advice on how to best go about it.

@@

We respectfully disagree that FAIR theory is like preregistration, and we think that this comment signals a need to better explain the important distinctions between these two practices.
To address this comment, we have included the following section in the revision:

@@

There are now three worked examples for actual theories in the package vignettes:

* [Self-Determination Theory](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html)
* The [Dunning-Kruger Effect](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html)
* [Morris' Tripartite Model](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html)

@@

That said, we are not sure that we are best positioned to engage with these issues, nor that the present paper would be the best place for such an argument.
As methodologists, our primary aim in this paper is to articulate the affordances of FAIR theory:
a framework that supports better theory specification, sharing, and reuse.
We believe we can make this contribution most effectively if we remain sharply focused on the practical infrastructure and its scientific use cases.

### 2025-06-19 test.Rmd
update 2025-06-19 09:59:24.514362

authornote: |
  We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion. We also thank the Editor, Dr. Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments. This version of the paper was generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
Applied to FAIR theory, reusability requires that  each theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and Accessible license for reuse (R1.1).
It should further have detailed provenance (R1.2), 
which is achieved through version control with Git and archival on Zenodo.
Finally, the (meta)data which meets domain-relevant community standards (R1.3).
The Datacite metadata schema offers an initial template in this regard,
and this paper takes one step towards establishing more fine-grained community standards for FAIR theory.
[This is an example of FAIR metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) extracted from Zenodo.

@@

When archiving a FAIR theory, documenting it with relevant metadata improves its Findability.
We recommend using a standardized metadata schema like DataCite [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
See here for [an example of the metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) associated with our FAIR empirical cycle.
Within this schema,

>>>>

authornote: |
  We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion. We also thank the Editor, Dr. Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments. Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #`r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
Applied to FAIR theory, reusability requires that  each theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and Accessible license for reuse (R1.1).
It should further have detailed provenance (R1.2), 
which is achieved through version control with Git and archival on Zenodo.
Finally, the (meta)data which meets domain-relevant community standards (R1.3).
The Datacite metadata schema offers an initial template in this regard,
and this paper takes one step towards establishing more fine-grained community standards for FAIR theory.
[This is an example of FAIR metadata](https://github.com/cjvanlissa/fair_theory/blob/3b4894da576cb76d19e911a05dd513d5172058ec/example_metadata.json) extracted from Zenodo.

@@

When archiving a FAIR theory, documenting it with relevant metadata improves its Findability.
We recommend using a standardized metadata schema like DataCite [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
See here for [an example of the metadata](https://github.com/cjvanlissa/fair_theory/blob/3b4894da576cb76d19e911a05dd513d5172058ec/example_metadata.json) associated with our FAIR empirical cycle.
Within this schema,
